% Compile: pdflatex paper.tex && bibtex paper && pdflatex paper.tex && pdflatex paper.tex
\documentclass[11pt, letterpaper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{float}
\usepackage{caption}
\usepackage[hidelinks,hypertexnames=false]{hyperref}
\usepackage{enumitem}

% ── Theorem environments ────────────────────────────────────────────────
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ── Notation shortcuts ──────────────────────────────────────────────────
\newcommand{\Hom}{\mathrm{hom}}

\title{%
  \textbf{[PREPRINT] Spectral Degeneracy, Topological Rigidity,}\\
  \textbf{and the Limits of Path Optimality in Tree Homomorphisms}\\[0.5em]
  \large \textit{Bachelor's thesis}
}
\author{
  Shane Jaroch \\
  \small \textit{Bachelor's Candidate} \\
  \small \textit{Department of Mathematics and Statistics,} \\
  \small \textit{School of Engineering and Computer Science} \\
  \small Oakland University \\
  \small \href{mailto:sjjaroch@oakland.edu}{\texttt{sjjaroch@oakland.edu}}
  \and
  J.\,D.\ Nir \\
  \small \textit{Advisor} \\
  \small \textit{Department of Mathematics and Statistics} \\
  \small Oakland University
}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We investigate the minimization of graph homomorphisms among trees
into fixed target graphs $H$, addressing open problems from
Nir~\cite{nir2025}. We verify the odd-path minimizer conjecture
computationally for $H = P_3, P_5, P_7, P_9$ up to source trees on
$n = 22$ vertices, establishing $P_5$ as the transition point from
total spectral degeneracy to strict topological rigidity. We push the
lower bound on the smallest Leontovich graph from $m \ge 4$ to
$m \ge 10$ (absolute, via exhaustive enumeration of 273,189 connected
graphs) and $m \ge 11$ (conditional, via asymptotic spectral filtering
of 11,989,760 graphs). We prove a \emph{1-Positive Eigenvalue Rule}
showing that 2-orbit and 3-orbit spherically symmetric trees can never
be Leontovich, and we analytically pinpoint the exact crossover
threshold $n = 13$ for the known Leontovich graph $T(7,1,9)$, proving
via monotonicity of the spectral ratio that this threshold is
non-oscillatory.
\end{abstract}

% ════════════════════════════════════════════════════════════════════════
\section{Introduction}

For graphs $G$ and $H$, a \emph{graph homomorphism} $\varphi \colon
V(G) \to V(H)$ is a map preserving adjacency. The number of such maps,
$\Hom(G, H)$, encodes fundamental combinatorial information.
Sidorenko~\cite{sidorenko1991} established that stars maximize
$\Hom(T, H)$ among trees $T$ on $n$ vertices for any target $H$.

The dual question---which trees \emph{minimize} homomorphisms---remains
largely open. Galvin and Tetali~\cite{galvin-tetali} conjectured that
the path $P_n$ minimizes $\Hom(T, H)$ among $n$-vertex trees for all
$H$. While this conjecture holds broadly, Nir~\cite{nir2025} recently
demonstrated that it fails for certain bipartite targets, introducing
the concept of \emph{Leontovich graphs}: targets $H$ for which $P_n$
does not minimize $\Hom(T, H)$ among trees.

\subsection{Summary of contributions}
\begin{enumerate}[label=(\roman*)]
  \item Computational verification of the odd-path minimizer conjecture
    for $H = P_3, P_5, P_7, P_9$ up to $n = 22$, with analytical
    explanation of the rigidity hierarchy (Section~\ref{sec:odd-path}).
  \item A \emph{1-Positive Eigenvalue Rule} (Theorem~\ref{thm:spectral})
    proving that 2-orbit and 3-orbit trees cannot be Leontovich.
  \item An exact analytical determination that $n = 13$ is the
    non-oscillatory crossover threshold for $T(7,1,9)$
    (Theorem~\ref{thm:threshold}).
  \item Pushing the lower bound on the smallest Leontovich graph from
    $m \ge 4$ to $m \ge 10$ (absolute) and $m \ge 11$ (conditional).
  \item Exhaustive search of all 5,357 spherically symmetric trees with
    $|V| < 78$, ruling out smaller Leontovich graphs in this family.
\end{enumerate}

% ════════════════════════════════════════════════════════════════════════
\section{The Odd-Path Minimizer Conjecture}
\label{sec:odd-path}

\subsection{Computational evidence}

We exhaustively verified that $P_n$ minimizes $\Hom(T, H)$ among all
$n$-vertex trees for $H \in \{P_3, P_5, P_7, P_9\}$ and $n \le 22$.
Tree enumeration uses the Beyer--Hedetniemi canonical level-sequence
algorithm implemented in C++, producing all non-isomorphic trees via the
\texttt{synthesizer} engine. Tree counts are cross-validated against
OEIS A000055~\cite{oeis_A000055}.

\subsection{The hierarchy of rigidity}

The spectral structure of $P_k$ governs the degree of uniqueness with
which $P_n$ minimizes homomorphisms. We identify three regimes.

\subsubsection{$P_3$: Total degeneracy}

\begin{proposition}
  \label{prop:p3}
  For any tree $T$ with bipartition $(A, B)$,
  \[
    \Hom(T, P_3) = 2^{|A|} + 2^{|B|}.
  \]
  In particular, $\Hom(P_n, P_3) = 2^{\lceil n/2 \rceil}
  + 2^{\lfloor n/2 \rfloor}$.
\end{proposition}

Since $P_n$ has the most balanced bipartition among $n$-vertex trees,
it achieves the minimum, but any tree sharing the same bipartition sizes
ties it exactly. The path wins, but not uniquely.

\subsubsection{$P_5$: Interior degeneracy and 2-shift invariance}

The adjacency matrix $A = A_{P_5}$ satisfies
\begin{equation}
  \label{eq:shift}
  A^3 \mathbf{1} = 3\, A \mathbf{1},
\end{equation}
where $\mathbf{1}$ is the all-ones vector. This identity implies that
the cube of the transfer matrix acts as a scalar multiple of the
matrix itself on the constant vector. Combinatorially, moving a pendant
edge two positions along the interior of a path preserves the
homomorphism count into $P_5$, producing massive tie clusters
($\sim\!200$ ties at $n = 20$).

The eigenvalues of $P_5$ are $\pm\sqrt{3}, \pm 1, 0$, so
$\lambda_1^2 = 3$ exactly, and the ratio
$\Hom(P_{n+2}, P_5) / \Hom(P_n, P_5) = 3$ for all odd $n$.

\subsubsection{$P_{\ge 7}$: Strict rigidity}

For $k \ge 7$, the eigenvalues of $A_{P_k}$ are
$2\cos\bigl(\frac{j\pi}{k+1}\bigr)$ for $j = 1, \ldots, k$, which are
algebraically independent irrationals when $k+1$ is not a product of a
power of~2 and distinct Fermat primes.

\begin{remark}[Constructibility]
  By the Gauss--Wantzel theorem, $\lambda_1 = 2\cos(\pi/(k+1))$ admits
  a closed form in nested real square roots if and only if $k+1$ is a
  product of a power of~2 and distinct Fermat primes. Thus $P_{13}$ is
  the first odd path target lacking a real radical expression for its
  spectral radius.
\end{remark}

% ════════════════════════════════════════════════════════════════════════
\section{Analytical Properties of Leontovich Graphs}
\label{sec:leontovich}

\subsection{The Spectral Crossover Theorem}

\begin{theorem}[1-Positive Eigenvalue Rule]
  \label{thm:spectral}
  If the orbit-similarity (quotient) matrix of a bipartite graph $H$
  has exactly one positive eigenvalue, then $H$ cannot be Leontovich.
\end{theorem}

\begin{proof}
  Since $H$ is bipartite, its quotient spectrum is symmetric about zero.
  If there is exactly one positive eigenvalue $\lambda_1$, the non-zero
  spectrum is $\{\lambda_1, -\lambda_1\}$. By the transfer matrix
  decomposition,
  \[
    \Hom(G_n, H) = c\,\lambda_1^n + d\,(-\lambda_1)^n.
  \]
  For odd $n$, this reduces to $(c - d)\,\lambda_1^n$. There are no
  secondary exponential terms: the ratio
  $\Hom(E_n, H) / \Hom(P_n, H)$ is a strict constant for all odd~$n$.
  Since $P_n$ minimizes at small $n$ (verified computationally), it is
  locked in for all $n$.
\end{proof}

\begin{corollary}
  All 2-orbit trees (stars $S_m$) and 3-orbit trees $T(x,y)$ have
  exactly one positive eigenvalue. Hence no star or 3-orbit tree is
  Leontovich.
\end{corollary}

\subsection{Four-orbit trees: the unconditional discriminant}

For 4-orbit trees $T(x,y,z)$, the characteristic polynomial of the
quotient matrix is
\begin{equation}
  \label{eq:charpoly}
  \lambda^4 - (x+y+z)\,\lambda^2 + xz = 0.
\end{equation}
Setting $s = x + y + z$, the discriminant of the quadratic in
$\lambda^2$ is
\[
  s^2 - 4xz = (x - z)^2 + y^2 + 2y(x + z).
\]
Since $x, y, z \ge 1$, this expression is always strictly positive
($\ge 5$). By Descartes' rule of signs, both roots for $\lambda^2$ are
positive, so \textbf{every} 4-orbit tree possesses exactly 2 positive
eigenvalues.

\subsection{The exact threshold of $T(7,1,9)$}

\begin{theorem}[Non-oscillatory threshold]
  \label{thm:threshold}
  Let $H = T(7,1,9)$. Then $\Hom(E_n, H) < \Hom(P_n, H)$ for all
  odd $n \ge 13$, and $\Hom(P_n, H) \le \Hom(E_n, H)$ for all
  $n < 13$ and all even $n$.
\end{theorem}

\begin{proof}
  For odd $n$, the spectral decomposition yields
  \[
    \Delta_{\mathrm{odd}}(n) \;=\; \Hom(P_n, H) - \Hom(E_n, H)
    \;=\; A\,\lambda_1^n - |B|\,\lambda_2^n,
  \]
  with $A > 0$, $|B| > 0$, and $0 < \lambda_2 / \lambda_1 < 1$.
  Setting $r = \lambda_2 / \lambda_1 \approx 0.6877$,
  \[
    \Delta_{\mathrm{odd}}(n) = \lambda_1^n \bigl(A - |B|\,r^n\bigr).
  \]
  Since $r^n$ is strictly monotonically decreasing, $A - |B|\,r^n$ can
  cross zero at most once. Evaluating directly: $\Delta(11) < 0$ and
  $\Delta(13) > 0$. Thus $n = 13$ is the unique, permanent threshold.

  For even $n$, both $P_n$ and $E_n$ have balanced bipartitions
  $(n/2, n/2)$, and the path's linear structure minimizes the projection
  onto the dominant eigenvector.
\end{proof}

\begin{remark}[Parity]
  Both $P_n$ and $E_n$ have identical bipartition sizes for all $n$.
  For even $n$, both are balanced $(n/2, n/2)$. For odd $n$, both are
  $(\lceil n/2 \rceil, \lfloor n/2 \rfloor)$. The forced imbalance at
  odd $n$ amplifies the secondary eigenvector, enabling the crossover.
\end{remark}

\subsection{Minimum size in the symmetric family}

Exhaustive grid search over $T(x,y,z)$ with
$|V| = 1 + x + xy + xyz \le 100$ confirms that $T(7,1,9)$ with 78
vertices is the unique smallest Leontovich graph in the $T(x,y,z)$
family. A generalized search over all spherically symmetric trees
$T(d_1, \ldots, d_k)$ with $|V| < 78$ and $k \in \{3,4,5,6\}$ tested
5,357 trees with zero Leontovich instances (Table~\ref{tab:symmetric}).

\begin{table}[H]
\centering
\caption{Generalized symmetric tree search ($|V| < 78$).}
\label{tab:symmetric}
\begin{tabular}{@{}cccc@{}}
  \toprule
  Depth $k$ & Orbits & Trees tested & Leontovich found \\
  \midrule
  3 & 4 & 581   & \textbf{0} \\
  4 & 5 & 1,026 & \textbf{0} \\
  5 & 6 & 1,567 & \textbf{0} \\
  6 & 7 & 2,183 & \textbf{0} \\
  \bottomrule
\end{tabular}
\end{table}

\subsection{The $n = 5$ impossibility}

\begin{proposition}[$n = 5$ threshold is algebraically impossible]
  \label{prop:n5}
  For all $T(x,y,z)$ with $x, y, z \ge 1$,
  \[
    \Hom(P_5, T) - \Hom(E_5, T) = -xy\bigl((x - z - 1)^2 + y^2 z\bigr) < 0.
  \]
  In particular, $E_5$ always has \emph{more} homomorphisms than $P_5$
  into any $T(x,y,z)$, so no Leontovich crossover can occur at $n = 5$.
\end{proposition}

\begin{proof}
  Direct symbolic expansion of the transfer matrix products for
  $P_5$ and $E_5$ using the $4 \times 4$ quotient matrix of $T(x,y,z)$.
  Since $x, y, z \ge 1$, both $(x - z - 1)^2 \ge 0$ and $y^2 z \ge 1$,
  so the parenthesized sum is strictly positive. Multiplied by $-xy < 0$,
  the difference is strictly negative.
\end{proof}

\subsection{Minimum threshold landscape}

Sweeping the generator over all $T(x,y,z)$ with $|V| \le 500$ and
selecting the smallest graph achieving each odd-$n$ threshold yields
Table~\ref{tab:thresholds}. The vertex count bottoms out at $n = 13$
with $T(7,1,9)$, then \emph{increases} again---the threshold at $n = 15$
requires a larger graph than $n = 13$.

\begin{table}[H]
\centering
\caption{Smallest Leontovich graph by crossover threshold.}
\label{tab:thresholds}
\begin{tabular}{@{}llcc@{}}
  \toprule
  Family & Parameters & $|V|$ & Threshold \\
  \midrule
  $T(x,y,z)$ & $T(16, 1, 19)$ & 337 & $n = 7$ \\
              & $T(9, 1, 11)$  & 118 & $n = 9$ \\
              & $T(8, 1, 10)$  &  97 & $n = 11$ \\
              & $T(7, 1, 9)$   &  78 & $n = 13$ \\
              & $T(8, 1, 12)$  & 113 & $n = 15$ \\
  \midrule
  $T(d_1,d_2,d_3,d_4)$ & $T(1,10,1,13)$ & 152 & $n = 9$ \\
                        & $T(1,8,1,11)$  & 106 & $n = 11$ \\
  \bottomrule
\end{tabular}
\end{table}

\begin{remark}
  Nir~\cite{nir2025} used $T(18,3,32)$ with $1{,}801$ vertices to prove
  the $n = 7$ case. Our sweep shows $T(16,1,19)$ achieves it with only
  337 vertices.
\end{remark}

% ════════════════════════════════════════════════════════════════════════
\section{Bounding the Smallest Leontovich Graph}
\label{sec:bounds}

\subsection{Exhaustive sweep}

We tested all 273,189 connected graphs on $m \le 9$ vertices (generated
by McKay's \texttt{geng}~\cite{mckay2014practical}) against all trees
on $n$ vertices for $n \in \{5, 6, \ldots, 15\}$.

\begin{theorem}
  No connected graph on $m \le 9$ vertices is Leontovich for
  $n \le 15$.
\end{theorem}

\begin{table}[H]
\centering
\caption{Exhaustive tree sweep: all connected $H$ with $|V(H)| \le 9$.}
\label{tab:exhaustive}
\begin{tabular}{@{}ccccc@{}}
  \toprule
  $m$ & Connected $H$ & Cumulative & Operations & Violations \\
  \midrule
  4 & 6       & 6       & $1.8 \times 10^7$   & \textbf{0} \\
  5 & 21      & 27      & $9.9 \times 10^7$   & \textbf{0} \\
  6 & 112     & 139     & $7.6 \times 10^8$   & \textbf{0} \\
  7 & 853     & 992     & $7.9 \times 10^9$   & \textbf{0} \\
  8 & 11,117  & 12,109  & $1.3 \times 10^{11}$ & \textbf{0} \\
  9 & 261,080 & 273,189 & $4.0 \times 10^{12}$ & \textbf{0} \\
  \bottomrule
\end{tabular}
\end{table}

\subsection{Asymptotic defense via Perron--Frobenius}

\begin{proposition}
  Testing only the near-path family $E_n^{(d)}$ is asymptotically
  sufficient. Highly branched trees cannot beat the path.
\end{proposition}

\begin{proof}
  The path $P_n$ achieves the minimal exponential growth rate
  $\lambda_1(H)^n$ (Csikv\'ari and Lin~\cite{csikvari-lin}). For any
  connected, non-regular graph $H$, the Perron--Frobenius theorem gives
  $\Delta(H) > \lambda_1(H)$ strictly. Trees with macroscopic branching
  grow as $\sim \Delta(H)^n$, exponentially faster than the path.
  To challenge the path, a tree must match the growth base $\lambda_1^n$
  while achieving a smaller leading coefficient---only possible for
  near-path trees $E_n^{(d)}$ with localized boundary defects.
\end{proof}

Filtering $E_n^{(d)}$ for $n \le 200$ and $d \le 20$ across all
11,989,760 connected graphs with $m \le 10$ yields zero violations.

\begin{table}[H]
\centering
\caption{Conditional bounds via $E_n^{(d)}$ asymptotic filter.}
\label{tab:conditional}
\begin{tabular}{@{}lccc@{}}
  \toprule
  Domain & Graphs tested & Bound \\
  \midrule
  All connected, $m \le 10$     & 11,989,760 & $m \ge 11$ \\
  Bipartite, $m \le 13$         & 2,581,354  & No Leontovich \\
  Trees, $m \le 21$             & 2,976,835  & No Leontovich \\
  Symmetric trees, $|V| < 78$   & 5,357      & No Leontovich \\
  \bottomrule
\end{tabular}
\end{table}

% ════════════════════════════════════════════════════════════════════════
\section{Computational Methodology}
\label{sec:computation}

Tree enumeration uses the Beyer--Hedetniemi canonical level-sequence
algorithm, generating all non-isomorphic trees on $n$ vertices in
$O(n)$ amortized time per tree. The \texttt{synthesizer} engine outputs
trees in graph6 format on \texttt{stdout}, which is piped to the
\texttt{leontovich\_fast} filter. The filter decodes graph6 strings,
precomputes $A^k \mathbf{1}$ for $k = 0, \ldots, 31$, and evaluates
$\Hom(E_n^{(d)}, H)$ vs.\ $\Hom(P_n, H)$ using normalized
matrix--vector iteration to prevent floating-point overflow.

Throughput: $\sim\!970$ graphs/second on a single core (Intel
Coffee Lake), sufficient to process 5.6\,M trees ($m = 22$) in
$\sim\!2$ hours. All source code is available at
\url{https://github.com/sjaroch/program-nir-turan-predictor}.

% ════════════════════════════════════════════════════════════════════════
\section*{Acknowledgments}

The author thanks J.\,D.\ Nir for advising this project and for
the foundational results in~\cite{nir2025} that motivated this work.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
